---
title: "Gov 51 - PSet 3"
author: "Tyler Dang"
format: pdf
editor: visual
---

# 1. Setup And Data Exploration

## Load Packages and Data

```{r}
library(tidyverse)
library(glmnet)
library(knitr)
```

```{r}
compas <- read.csv("data/raw/compas_clean.csv")
```

## Observations and Variables? Unit?

```{r}
dim(compas)
```

There are 7214 observations in the Compas dataset and 17 variables.

## Examine the Outcome Variable

```{r}
mean(compas$recidivism)
```

The recidivism rate is .451.

The outcome is binary as it takes only two values: 0 or 1.

This matters for linear models because a linear probability model produces predictions as probabilities of recidivism, but linear models can produce predictions below 0 or above 1 which are not reasonable as probabilities, but rather a change in probability.

## Summary Statistics

```{r}
summary_table <- data.frame(variable = names(compas),
                            mean = sapply(compas, mean, na.rm = TRUE),
                            sd   = sapply(compas, sd, na.rm = TRUE))
kable(summary_table, digits = 3)
```

We can see that age with a sd of 11.889 has the largest variation, while native_american (the dummy variable for whether a observation is native american) a sd of .050 has the least variation.

## Explore Key Predictors

```{r}
compas$priors_cat <- NA

compas$priors_cat[compas$priors_count == 0] <- "0"
compas$priors_cat[compas$priors_count %in% 1:2] <- "1-2"
compas$priors_cat[compas$priors_count %in% 3:5] <- "3-5"
compas$priors_cat[compas$priors_count %in% 6:10] <- "6-10"
compas$priors_cat[compas$priors_count >= 11] <- "11+"

compas$priors_cat <- factor(compas$priors_cat,
                            levels = c("0", "1-2", "3-5", "6-10", "11+"))

rates <- compas %>%
  group_by(priors_cat) %>%
  summarise(rate = mean(recidivism, na.rm = TRUE))

ggplot(rates, aes(x = priors_cat, y = rate, fill = priors_cat)) +
  geom_bar(stat = "identity") +
  labs(x = "Number of Prior Offenses",
       y = "Recidivism Rate") + 
  theme(legend.position = "none")
```

We can see that as the number of prior offenses increases, the recidivism rate increases.

# 2. OLS Baseline

## Train/test Split

## Fit OLS with Main Effects Only

## In-Sample vs Out-of-sample RMSE

## Cross-Validation by Hand

# 3. The Kitchen Sink: What Happens when You Add Too Many Predictors?

## Build the Expanded Model

## Fit OLS with the Kitchen Sink

## The Overfitting Lesson

# 4. Ridge Regression: Shrink Everything

## Fit Ridge with Cross-validation

## Ridge RMSE

## Did Ridge set any Coefficients to Zero

# 5. LASSO Regression: Shrink and Select

## Fit LASSO with Cross–Validation

## LASSO Variable Selection

## LASSO RMSE

# 6. Model Comparison and Interpretation

## Elastic Net

## Big Comparison Table

## Baker Article Connection

## Prediction vs Explanation

## Algorithmic Fairness
