---
title: "Gov 51 - PSet 3"
author: "Tyler Dang"
format: pdf
editor: visual
---

# 1. Setup And Data Exploration

## Load Packages and Data

```{r}
library(tidyverse)
library(glmnet)
library(knitr)
```

```{r}
compas <- read.csv("data/raw/compas_clean.csv")
```

## Observations and Variables? Unit?

```{r}
dim(compas)
```

There are 7214 observations in the Compas dataset and 17 variables.

## Examine the Outcome Variable

```{r}
mean(compas$recidivism)
```

The recidivism rate is .451.

The outcome is binary as it takes only two values: 0 or 1.

This matters for linear models because a linear probability model produces predictions as probabilities of recidivism, but linear models can produce predictions below 0 or above 1 which are not reasonable as probabilities, but rather a change in probability.

## Summary Statistics

```{r}
summary_table <- data.frame(variable = names(compas),
                            mean = sapply(compas, mean, na.rm = TRUE),
                            sd   = sapply(compas, sd, na.rm = TRUE))
kable(summary_table, digits = 3)
```

We can see that age with a sd of 11.889 has the largest variation, while native_american (the dummy variable for whether a observation is native american) a sd of .050 has the least variation.

## Explore Key Predictors

```{r}
compas$priors_cat <- NA

compas$priors_cat[compas$priors_count == 0] <- "0"
compas$priors_cat[compas$priors_count %in% 1:2] <- "1-2"
compas$priors_cat[compas$priors_count %in% 3:5] <- "3-5"
compas$priors_cat[compas$priors_count %in% 6:10] <- "6-10"
compas$priors_cat[compas$priors_count >= 11] <- "11+"

compas$priors_cat <- factor(compas$priors_cat,
                            levels = c("0", "1-2", "3-5", "6-10", "11+"))

rates <- compas %>%
  group_by(priors_cat) %>%
  summarise(rate = mean(recidivism, na.rm = TRUE))

ggplot(rates, aes(x = priors_cat, y = rate, fill = priors_cat)) +
  geom_bar(stat = "identity") +
  labs(x = "Number of Prior Offenses",
       y = "Recidivism Rate") + 
  theme(legend.position = "none")
```

We can see that as the number of prior offenses increases, the recidivism rate increases.

# 2. OLS Baseline

## Train/test Split

```{r}
set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]

nrow(train)
nrow(test)
```

There are 5771 observations in the training set, and 1443 in the test set.

## Fit OLS with Main Effects Only

```{r}
set.seed(51)
train_nocat <- subset(train, select = -priors_cat)
model2.2 <- lm(recidivism ~ ., data = train_nocat)
summary(model2.2)
```

Variables with a significant at the 5% level: age, female, priors_count, felony, age_25_45, has_juv_misd.

A one-unit increase in the number of prior offenses is associated with a 0.0295 increase in the probability of recidivism, holding all other variables constant.

## In-Sample vs Out-of-sample RMSE

```{r}
set.seed(51)
yhat_train <- predict(model2.2, newdata = train)
rmse_train <- sqrt(mean((yhat_train - train$recidivism)^2))
rmse_train

yhat_test <- predict(model2.2, newdata = test)
rmse_test <- sqrt(mean((yhat_test - test$recidivism)^2))
rmse_test

rmse_test - rmse_train
```

The in-sample RMSE: 0.4606

The out-of-sample RMSE: 0.4622

The gap is 0.0016, which is very small, indicating that the model is nearly identical for the training and test data, so there is no meaningful overfitting with 16 predictors.

## Cross-Validation by Hand

```{r}
set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train_nocat)))
rmse_cv <- numeric(5)

for (k in 1:5) {
  train_k <- train_nocat[folds != k, ]
  test_k <- train_nocat[folds == k, ]
  
  model_k <- lm(recidivism ~ ., data = train_k)
  pred_k <- predict(model_k, newdata = test_k)
  
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))}

mean(rmse_cv)
```

A 5-fold CV RMSE of 0.4620 is extremely close to the single-split test RMSE, indicating very similar estimated out-of-sample performance.

Cross-validation is more reliable because it averages over multiple splits instead of using a single split that may not be representative. Using multiple splits therefore will reduce variability and give more stable estimates.

Cross-validation is correct for choosing $\lambda$ because it minimizes prediction error. By evaluating many possible values, it balances bias and variance thus reducing overfitting.

# 3. The Kitchen Sink: What Happens when You Add Too Many Predictors?

## Build the Expanded Model

```{r}
train_nocat <- subset(train, select = -priors_cat)
test_nocat  <- subset(test,  select = -priors_cat)

f_full <- recidivism ~ (.)^2 +
  I(age^2) + I(priors_count^2) +
  I(juv_fel_count^2) + I(juv_misd_count^2) +
  I(juv_other_count^2)

X_train <- model.matrix(f_full, data = train_nocat)[, -1]
X_test  <- model.matrix(f_full, data = test_nocat)[, -1]

y_train <- train_nocat$recidivism
y_test  <- test_nocat$recidivism

ncol(X_train)
nrow(train_nocat)
nrow(train_nocat) / ncol(X_train)
```

The expanded model has 141 predictors.

Thus the ratio is 40.9 which means there are less observations per predictor increasing the chance of overfitting.

## Fit OLS with the Kitchen Sink

```{r}
ols_full <- lm(f_full, data = train_nocat)

pred_train <- predict(ols_full, newdata = train_nocat)
rmse_train <- sqrt(mean((train_nocat$recidivism - pred_train)^2))
rmse_train

pred_test <- predict(ols_full, newdata = test_nocat)
rmse_test <- sqrt(mean((test_nocat$recidivism - pred_test)^2))
rmse_test

rmse_test - rmse_train
```

The in-sample RMSE is 0.4485

The out-of-sample RMSE is 0.4890

The difference is 0.0405

Compared to the simple model with a gap of about 0.0016, this gap is much larger. This expanded model is overfitting. With so many predictors, the model fits the training data better, which is why there is a lower in-sample RMSE, but performs worse on new data, leading to a higher out-of-sample RMSE. This happens because the model is capturing noise rather than true underlying patterns.

## The Overfitting Lesson

Adding so many interaction terms increased the models flexibility, so it began to individually fit every data point instead of finding a good fit. This led to increased noise in the model. When we use a new data set, the model did not predict well because it was still using the noise from the training set. This is overfitting.

# 4. Ridge Regression: Shrink Everything

## Fit Ridge with Cross-validation

```{r}
set.seed(51) # for reproducible CV folds
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)

cv_ridge$lambda.min
plot(cv_ridge)
```

The optimal lambda is 0.0138

From the graph we can see that prediction error decreases as $\lambda$ increases until it reaches an optimal balance at the minimum before error flattens.

## Ridge RMSE

```{r}
pred_ridge <- predict(cv_ridge, s = "lambda.min", newx = X_test)
rmse_ridge <- sqrt(mean((y_test - pred_ridge)^2))
rmse_ridge
```

Simple OLS RMSE: 0.4622

Kitchen-Sink OLS RMSE: 0.489

Ridge RMSE: 0.470

The ridge regression improved to limit the overfitting from kitchen-sink by shrinking the coefficients thus reducing variance. It does not outperform the simple model though which indicates that the increase in interaction terms does not necessarily help.

## Did Ridge set any Coefficients to Zero

```{r}
coef(cv_ridge, s = "lambda.min")
```

No, Ridge did not eliminate any variables. Ridge shrinks coefficients towards zero but does not set them exactly to zero. So, we can see that most coefficients are very small but not zero. The entries shown as "." are not zero.

# 5. LASSO Regression: Shrink and Select

## Fit LASSO with Cross–Validation

```{r}
set.seed(51)
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)

cv_lasso$lambda.min

plot(cv_lasso)
```

The optimal $\lambda$ is 0.00769

## LASSO Variable Selection

```{r}
coef_lasso <- coef(cv_lasso, s = "lambda.min")

nonzero <- sum(coef_lasso != 0)
zero    <- sum(coef_lasso == 0)

nonzero
zero
```

Excluding the "intercept" there are 27 variables retained (the nonzero minus one). Thus, there are 114 eliminated.

```{r}
coef_mat <- as.matrix(coef_lasso)
nonzero_vars <- rownames(coef_mat)[coef_mat != 0]
nonzero_vars
```

Some surviving interaction terms: age:female, female:priors_counts, and many more. These make substantive sense as they suggest that the effect of one risk factor may depend on other characteristic rather than being the same for all.

LASSO reduces this subjective discretion because it selects variables on predictive performance than some choice of the researcher. So, instead of a human using their own intuition or biases to decide which interactions to include or not include, LASSO keeps the interactions that improve out-of sample prediction

## LASSO RMSE

```{r}
pred_lasso <- predict(cv_lasso, s = "lambda.min", newx = X_test)

rmse_lasso <- sqrt(mean((y_test - pred_lasso)^2))
rmse_lasso
```

Simple OLS RMSE: 0.4622

Kitchen-Sink OLS RMSE: 0.489

Ridge RMSE: 0.470

LASSO RMSE: 0.4593

Lasso performs the best for out-of-sample RMSE for all tests we have done.

# 6. Model Comparison and Interpretation

## Elastic Net

## Big Comparison Table

## Baker Article Connection

## Prediction vs Explanation

## Algorithmic Fairness
