---
title: "Gov 51 - PSet 3"
author: "Tyler Dang"
format: pdf
editor: visual
---

# 1. Setup And Data Exploration

## Load Packages and Data

```{r}
library(tidyverse)
library(glmnet)
library(knitr)
```

```{r}
compas <- read.csv("data/raw/compas_clean.csv")
```

## Observations and Variables? Unit?

```{r}
dim(compas)
```

There are 7214 observations in the Compas dataset and 17 variables.

## Examine the Outcome Variable

```{r}
mean(compas$recidivism)
```

The recidivism rate is .451.

The outcome is binary as it takes only two values: 0 or 1.

This matters for linear models because a linear probability model produces predictions as probabilities of recidivism, but linear models can produce predictions below 0 or above 1 which are not reasonable as probabilities, but rather a change in probability.

## Summary Statistics

```{r}
summary_table <- data.frame(variable = names(compas),
                            mean = sapply(compas, mean, na.rm = TRUE),
                            sd   = sapply(compas, sd, na.rm = TRUE))
kable(summary_table, digits = 3)
```

We can see that age with a sd of 11.889 has the largest variation, while native_american (the dummy variable for whether a observation is native american) a sd of .050 has the least variation.

## Explore Key Predictors

```{r}
compas$priors_cat <- NA

compas$priors_cat[compas$priors_count == 0] <- "0"
compas$priors_cat[compas$priors_count %in% 1:2] <- "1-2"
compas$priors_cat[compas$priors_count %in% 3:5] <- "3-5"
compas$priors_cat[compas$priors_count %in% 6:10] <- "6-10"
compas$priors_cat[compas$priors_count >= 11] <- "11+"

compas$priors_cat <- factor(compas$priors_cat,
                            levels = c("0", "1-2", "3-5", "6-10", "11+"))

rates <- compas %>%
  group_by(priors_cat) %>%
  summarise(rate = mean(recidivism, na.rm = TRUE))

ggplot(rates, aes(x = priors_cat, y = rate, fill = priors_cat)) +
  geom_bar(stat = "identity") +
  labs(x = "Number of Prior Offenses",
       y = "Recidivism Rate") + 
  theme(legend.position = "none")
```

We can see that as the number of prior offenses increases, the recidivism rate increases.

# 2. OLS Baseline

## Train/test Split

```{r}
set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]

nrow(train)
nrow(test)
```

There are 5771 observations in the training set, and 1443 in the test set.

## Fit OLS with Main Effects Only

```{r}
set.seed(51)
train_nocat <- subset(train, select = -priors_cat)
model2.2 <- lm(recidivism ~ ., data = train_nocat)
summary(model2.2)
```

Variables with a significant at the 5% level: age, female, priors_count, felony, age_25_45, has_juv_misd.

A one-unit increase in the number of prior offenses is associated with a 0.0295 increase in the probability of recidivism, holding all other variables constant.

## In-Sample vs Out-of-sample RMSE

```{r}
set.seed(51)
yhat_train <- predict(model2.2, newdata = train)
rmse_train <- sqrt(mean((yhat_train - train$recidivism)^2))
rmse_train

yhat_test <- predict(model2.2, newdata = test)
rmse_test <- sqrt(mean((yhat_test - test$recidivism)^2))
rmse_test

rmse_test - rmse_train
```

The in-sample RMSE: 0.4606

The out-of-sample RMSE: 0.4622

The gap is 0.0016, which is very small, indicating that the model is nearly identical for the training and test data, so there is no meaningful overfitting with 16 predictors.

## Cross-Validation by Hand

```{r}
set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train_nocat)))
rmse_cv <- numeric(5)

for (k in 1:5) {
  train_k <- train_nocat[folds != k, ]
  test_k <- train_nocat[folds == k, ]
  
  model_k <- lm(recidivism ~ ., data = train_k)
  pred_k <- predict(model_k, newdata = test_k)
  
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))}

mean(rmse_cv)
```

A 5-fold CV RMSE of 0.4620 is extremely close to the single-split test RMSE, indicating very similar estimated out-of-sample performance.

Cross-validation is more reliable because it averages over multiple splits instead of using a single split that may not be representative. Using multiple splits therefore will reduce variability and give more stable estimates.

Cross-validation is correct for choosing $\lambda$ because it minimizes prediction error. By evaluating many possible values, it balances bias and variance thus reducing overfitting.

# 3. The Kitchen Sink: What Happens when You Add Too Many Predictors?

## Build the Expanded Model

## Fit OLS with the Kitchen Sink

## The Overfitting Lesson

# 4. Ridge Regression: Shrink Everything

## Fit Ridge with Cross-validation

## Ridge RMSE

## Did Ridge set any Coefficients to Zero

# 5. LASSO Regression: Shrink and Select

## Fit LASSO with Cross–Validation

## LASSO Variable Selection

## LASSO RMSE

# 6. Model Comparison and Interpretation

## Elastic Net

## Big Comparison Table

## Baker Article Connection

## Prediction vs Explanation

## Algorithmic Fairness
